---
title: "Tutorial 1, Our first model."
author: "Arun Oakley-Cogan"
date: "2024-09-04"
output: html_document
---

```{r message = F}
library(tidyverse)
library(rethinking)
```

### Our first model. Estimating the presence/absence of frogs at a pond.

**Scenario**: We go to a pond 30 times and record the presence (1) or absence (0) of frogs, and estimate the probability of seeing a frog when visiting the pond.

**Workflow**

1.  Define what we are estimating
2.  Build generative model to simulate data
3.  Build statistical model
4.  Validate statistical model with simulated data
5.  Apply to real data.

### Generative model and simulated data.

```{r}
# number of trips to pond
n_trips <- 30

# the true probability of seeing a frog in this pond
true_prob <- 0.3

# simulate data from generative model
sim_data <- rbinom(n_trips, size = 1, prob = true_prob)

# data for model
data_list <- list(
  trips = length(sim_data), # total number of trips
  frogs = sum(sim_data) # number of times seen frogs
)
sim_data;
data_list
```

### Build statistical model: grid approximation

**Model Components**

-   **Grid**, a finite set of parameters values to be used
-   **Prior Distribution**, prior probability for each value of the parameter
-   **Data Likelihood**, the distribution function assigned to our observed variables. This function gives us the likelihood of the data for every plausible value of our parameter.
-   **Posterior Distribution**, contains the probability of different parameter values, conditional on the data and model.

#### Bayesian Notation

$$F \sim Binomial(n, p)$$ $$p \sim Uniform(0,1)$$

```{r}
# grid approximation method
# define our parameter grid
par_grid <- seq(from=0, to=1, length.out=20)

# define a uniform prior
prior <- rep(1, length(par_grid))
prior <- prior / sum(prior)

# estimate the likelihood of seeing frogs in this many trips
likelihood <- dbinom(data_list$frogs, data_list$trips, prob = par_grid)

# estimate posterior
posterior <- (likelihood * prior) / sum(likelihood * prior)

# plot posterior distribution
plot(par_grid, posterior, col='blue', type='b' , xlab='Probability of seeing frogs', ylab='Density')
# plot prior distribution
points(x = par_grid, y = prior, type='b', col='orange')
abline(v = 0.3, col="red")
legend("topright", legend=c("Posterior", "Prior", "True Probability"),
       col=c("blue", "orange", "red"), lty=1)

```

```{r}
# summarise our posterior - grid approximation
samples <- sample( par_grid , prob=posterior , size=1000 , replace=TRUE )

# mean
mean(samples)

# how much of the posterior is above/below a value?
sum( posterior[ par_grid < 0.4 ] )
sum( posterior[ par_grid > 0.4 ] )

# compatibility intervals
# percentile intervals (PI) - even distribution above/below interval, indicate shape of distribution
PI(samples, prob=0.5) 
# highest posterior density interval (HPDI) - best representatives of parameter values
HPDI(samples, prob=0.5)

```

### Build statistical model: MCMC (ulam)

```         
a.  **Model data,** a list containing our variables and data
b.  **Prior Distribution**, prior probability of the parameter at each probability
c.  **Data Likelihood**, the distribution function assigned to our observed variables. This function gives us the likelihood of the data for every plausible value of our parameter.
```

```{r}
# model data
data_list
```

```{r}
# MCMC method using ulam
frog_model <- ulam(
  alist(
    frogs ~ dbinom(trips, p), # data likelihood
    p ~ dunif(0, 1) # p prior
  ), data = data_list
)

```

```{r}
# model summary
precis(frog_model)
```

```{r}
# posterior samples
samples <- extract.samples(frog_model, n=10000)

# plots
curve(dnorm(x, mean(samples$p), sd(samples$p)), xlab="Probability of seeing frogs", ylab="Density", col="blue")
curve((dunif(x, 0, 1)), add = TRUE, col="orange")
abline(v = 0.3, col="red")
legend("topright", legend=c("Posterior", "Prior", "True Probability"),
       col=c("blue", "orange", "red"), lty=1)
```

```{r}
# posterior summaries
# above 30%
sum(samples$p > 0.3) / length(samples$p)

# compatibility intervals
# percentage interval
PI(samples$p, prob=0.5)
# highest posterior density interval
HPDI(samples$p, prob=0.5)
```
