---
title: "Tutorial 6 - Binomial Regression"
author: "Arun Oakley-Cogan"
date: "2024-10-21"
output: html_document
---

Today, we are working through the first in our series in Generalized Linear Models (glms), the most useful glms deal with counts (non-negative integers). Count data can be difficult to predict with our models as the scale of our predictors are never the same as our outcome scale

**Models for Count Data**

-   Binomial regression - used when data is binary, present/absent, 0/1. 
-   Poisson regression - used when count data has a unknown maximum.

Binomial regressions are also called **logistic regression** when data is in single-trial cases,\
or an **aggregated binomial regression** when the outcome is from 0 -\> n (number of total trials).

```{r}
library(rethinking)
library(ggplot2)
rm(list=ls())
```

We are going to be looking at a data set of the Southern Corroboree frogs that details the presence/absence at a number of different pools, in different sites in the Snowy Mountains, in Australia. Each row in the data is a unique pool in a particular site.

```{r}
fdata <- read.csv2('Data/frogs.csv')
precis(fdata)
```

**Variables:**

1.  pres.abs – 0/1 indicates whether frogs were found.
2.  easting – reference point
3.  northing – reference point
4.  altitude – altitude in meters
5.  distance – distance in meters to nearest extant population
6.  NoOfPools– number of potential breeding pools
7.  NoOfSites– number of potential breeding sites within a 2 km radius
8.  avrain – mean rainfall for Spring period
9.  meanmin – mean minimum Spring temperature
10. meanmix - mean maximum Spring temperature

One way to visualize this data is to plot the pools by their spatial relationship.

```{r}
plot(northing ~ easting, data = fdata, pch = c(1, 16)[fdata$pres.abs +
1], col=col.alpha("deepskyblue4", .7) ,xlab = "Meters east of reference point", ylab = "Meters north")
```

### Define what we are estimating

**Q1. Predict the probability of finding a frog in a pool**

**Q2**. **Understand the relationship between the occurrence of frogs and other environment variables**

This falls under a binomial (logistic) regression as our outcome variable pres.abs is binary 0/1.

```{r}
plot(pres.abs ~ avrain, data = fdata, pch = c(1, 16)[fdata$pres.abs +
1], col=col.alpha("deepskyblue4", .7) ,xlab = "Average Spring Rainfall (mm)", ylab = "Presence/Absence")
```

### Q1 Statistical model

$$
\large{
\begin{align*}
Pressence \sim Binomial(1, p)\\
p = invlogit(alpha)\\
alpha \sim Normal(?,?)\\
\end{align*}}
$$

```{r}
model_data <- list(
  pressence = fdata$pres.abs,
  avgrain = standardize(fdata$avrain),
  distance = standardize(fdata$distance),
  pools = standardize(fdata$NoOfPools)
)
```

```{r}
frog_model <- ulam(
  alist(
    pressence ~ dbinom(1, p),
    logit(p) <- alpha,
    alpha ~ dnorm(0, 1.5)
  ), data = model_data, chains = 4, log_lik = T
)
```

```{r}
precis(frog_model)
```

```{r}
prior1 <- extract.prior(frog_model)
par(mfrow=c(1,2))
plot(density(prior1$alpha), main="Prior prediction log-odds scale", col="deepskyblue4", xlab="Probability of occurence")
plot(density(inv_logit(prior1$alpha)),  main="Prior prediction outcome scale", col="deepskyblue4", xlab="Probability of occurence")
```

I have no prior knowledge of what the probability of seeing a frog should be, so I want to make this as flat as possible. Let's try another prior with a standard deviation of 1.6.

```{r}
frog_model <- ulam(
  alist(
    pressence ~ dbinom(1, p),
    logit(p) <- alpha,
    alpha ~ dnorm(0, 1.6)
  ), data = model_data, chains = 4, log_lik = T
)

prior1 <- extract.prior(frog_model)
par(mfrow=c(1,2))
plot(density(prior1$alpha), main="Prior prediction log-odds scale", col="deepskyblue4", xlab="Probability of occurence")
plot(density(inv_logit(prior1$alpha)),  main="Prior prediction outcome scale", col="deepskyblue4", xlab="Probability of occurence")
```

Not much of a change, but at least a little flatter. Let's go with this for now.

#### Convergence diagnostics

```{r}
# r_hat and ess
precis(frog_model)
```

```{r, fig.width = 15, fig.height = 5}
# looking for healthy caterpillars
# Trace plot
# stationary, path of a chain stays within the same region of the posterior
# good mixing, chains rapidly explore the full region, it doesn't slowly wander
# convergence, multiple, independent chains stick stick around the same region
traceplot(frog_model)
```

Happy with our convergence and priors, onto our inference!

#### Posterior Plots

```{r}
posterior_p <- link(frog_model)
plot(density(posterior_p),  main="Posterior prediction (outcome)", col="deepskyblue4", xlab="Probability of occurence", xlim=c(0,1))
lines(density(inv_logit(prior1$alpha)), col="orange")
legend("topleft", legend=c('Posterior', 'Prior'), col=c("deepskyblue4", "orange"), lty=1)
abline(v=mean(posterior_p))
```

Lets add some more predictors

### Statistical model

$$
\large{
\begin{align*}
Pressence \sim Binomial(1, p)\\
p = invlogit(alpha + betaR*avrain + betaP*pools + betaD*distance)\\
alpha \sim Normal(0,1.6)\\
beta R \sim Normal(?,?)\\
beta P \sim Normal(?,?)\\
beta D \sim Normal(?,?)\\
\end{align*}}
$$

What priors should we start with here for our predictors?? Our beta parameters here are representing the change in the log odds of finding a frog when each have been increased by 1. For me, I am not great at thinking about log odds so I will try some flat priors here as well.

```{r}
frog_full_model <- ulam(
  alist(
    pressence ~ dbinom(1, p),
    logit(p) <- alpha + betaR*avgrain + betaD*distance + betaP*pools,
    alpha ~ dnorm(0,1.6),
    betaR ~ dnorm(0, 1),
    betaD ~ dnorm(0, 1),
    betaP ~ dnorm(0, 1)
  ), data = model_data, chains = 4, log_lik = T
)
```

#### Prior predictive

```{r}
prior2 <- extract.prior(frog_full_model)
prior_p <- link(frog_full_model, post=prior2)
plot(density(prior_p), main="Prior prediction (outcome)", col="deepskyblue4", xlab="Probability of occurence")
```

We are getting a bit of a horns effect here, with the densities piling up closer to 0 and 1. Lets tighten these priors a little.

```{r}
frog_full_model <- ulam(
  alist(
    pressence ~ dbinom(1, p),
    logit(p) <- alpha + betaR*avgrain + betaD*distance + betaP*pools,
    alpha ~ dnorm(0,1.6),
    betaR ~ dnorm(0, 0.5),
    betaD ~ dnorm(0, 0.5),
    betaP ~ dnorm(0, 0.5)
  ), data = model_data, chains = 4, log_lik = T
)
prior2 <- extract.prior(frog_full_model)
prior_p <- link(frog_full_model, post=prior2)
plot(density(prior_p), main="Prior prediction (outcome)", col="deepskyblue4", xlab="Probability of occurence")
```

This is better, we have a fairly flat prior here. let's run with this and check our convergence.

#### Convergence diagnostics

```{r}
precis(frog_full_model)
traceplot(frog_full_model, pars=c('alpha', 'betaR'))
```

```{r}
plot(coeftab(frog_full_model))
```

Interpretation of coefficients.

-   Number of pools, mean: 0.33

The log odds of finding a frog increases by 0.32 when the number of pools increases by 1. Or the odds of finding a frog get multiplied by exp(0.32) (1.39) when increasing the number of pools by 1.

#### Posterior predictive

```{r, fig.width = 15, fig.height = 5}
# observed vs predicted
# make predictions of p from our model
post_p <- link(frog_full_model)
mean_p <- apply(post_p, 2, mean)
ci_p <- apply(post_p, 2, PI)


# get pools with frogs but low probability
pres_low <- model_data$pressence == 1 & mean_p < 0.3
# get pools with no frogs but high probability
pres_high <- model_data$pressence == 0 & mean_p > 0.6

par(mfrow=c(1,2))
plot(northing ~ easting, data = fdata, pch = c(1, 16)[fdata$pres.abs +
1], col=c("deepskyblue4","orange")[factor(pres_low)] ,xlab = "Meters east of reference point", ylab = "Meters north", main="Pools with no frogs but high prob")

plot(northing ~ easting, data = fdata, pch = c(1, 16)[fdata$pres.abs +
1], col=c("deepskyblue4","red")[factor(pres_high)] ,xlab = "Meters east of reference point", ylab = "Meters north", main="Pools with frogs but low prob")
```

```{r}
# another visualization of the posterior estimates
df <- data.frame(occ=factor(fdata$pres.abs), prob=mean_p)
# Basic violin plot
p <- ggplot(df, aes(x=occ, y=mean_p)) + 
  geom_violin() 
# violin plot with mean points
p + stat_summary(fun.y=mean, geom="point", shape=23, size=2) + geom_boxplot(width=0.1)

```

#### Model comparison

```{r}
compare(frog_model, frog_full_model)
```
