---
title: "Tutorial 1, Our first model."
author: "Arun Oakley-Cogan"
date: "2024-09-04"
output: html_document
---

```{r message = F}
library(tidyverse)
library(rethinking)
```

### Our first model. Estimating the presence/absence of frogs at a pond.

**Scenario**: We go to a pond 30 times and record the presence (1) or absence (0) of frogs, and estimate the probability of seeing a frog when visiting the pond.

**Workflow**

1.  Define what we are estimating - p, the prob of seeing a frog
2.  Build generative model to simulate data
3.  Build statistical model
4.  Validate statistical model with simulated data
5.  Apply to real data.

### Generative model and simulated data.

```{r}
# number of trips to pond

# the true probability of seeing a frog in this pond

# simulate data from generative model

# data for model

```

### Build statistical model: grid approximation

**Model Components**

-   **Grid**, a finite set of parameters values to be used
-   **Prior Distribution**, prior probability for each value of the parameter
-   **Data Likelihood**, the distribution function assigned to our observed variables. This function gives us the likelihood of the data for every plausible value of our parameter.
-   **Posterior Distribution**, contains the probability of different parameter values, conditional on the data and model.

#### Bayesian Notation

$$F \sim Binomial(n, p)$$ $$p \sim Uniform(0,1)$$

```{r}
# grid approximation method
# define our paramatrer grid


# define a uniform prior

# estimate the data likelihood of seeing frogs this many times in trips

# estimate posterior

# plot posterior distribution
#plot(par_grid, posterior, col='blue', type='b' , xlab='Probability of seeing frogs', ylab='Density')
# plot prior distribution
#points(x = par_grid, y = prior, type='b', col='orange')
#abline(v = true_prob, col="red")
#legend("topright", legend=c("Posterior", "Prior", "True Probability"), col=c("blue", "orange", "red"), lty=1)

```

```{r}
# summarise our posterior - grid approximation sample(x,p,size,rep)
# mean

# how much of the posterior is above/below a value? x[p<value]

# compatibility intervals
# highest posterior density interval (HPDI) - best representatives of parameter values

```

3.  Build our model: **MCMC (ulam)**
    a.  **Model data,** a list containing our variables and data
    b.  **Prior Distribution**, prior probability of the parameter at each probability
    c.  **Data Likelihood**, the distribution function assigned to our observed variables. This function gives us the likelihood of the data for every plausible value of our parameter.

```{r}
# model data
```

```{r}
# MCMC method using ulam
```

```{r}
# model summary
```

```{r}
# posterior samples
samples <- extract.samples(frog_model, n = 10000)
# plots
#curve(dnorm(x, mean(samples$p), sd(samples$p)), xlab="Probability of seeing frogs", ylab="Density", col="blue")
#curve((dunif(x, 0, 1)), add = TRUE, col="orange")
#abline(v = true_prob, col="red")
#legend("topright", legend=c("Posterior", "Prior", "True Probability"), col=c("blue", "orange", "red"), lty=1)
```

```{r}
# posterior summaries
# above 30%

# compatibility intervals
# highest posterior density interval

```
