---
title: "Tutorial 0, Software Installation for Rethinking course"
author: "Arun Oakley-Cogan"
date: "2024-08-28"
output: html_document
---

### 1. Install C++ Toolchain

Do you need it?? Open up command prompt (Windows) or Terminal (macOS)

#### macOS

clang++ --version
make --version

#### Windows

g++ --version
make --version

If you get a response from both of these commands you can go to step 2! Otherwise,

Go to: <https://mc-stan.org/docs/cmdstan-guide/cmdstan-installation.html#cpp-toolchain>

and follow instructions for your operating system

### 2. Install CmdStanR

```{r}
install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
cmdstanr::install_cmdstan()
```

### 3. Install Rethinking Package

```{r}
install.packages(c("coda","mvtnorm","devtools","loo","dagitty","shape"))
devtools::install_github("rmcelreath/rethinking")
```

### 4. Install Common R packages

```{r}
install.packages(c("ggplot2", "tidyverse", "bayesplot"))
```

### 5. Test Installation

```{r}
library(rethinking)

fit_stan <- ulam(
  alist(
    y ~ dnorm( mu , sigma ),
    mu ~ dnorm( 0 , 10 ),
    sigma ~ dexp( 1 )
  ), data=list(y=c(-1,1)) 
)
precis(fit_stan)
```
