---
title: "Tutorial 3 - All about the DAGS"
author: "Arun Oakley-Cogan"
date: "2024-09-26"
output: html_document
---

```{r}
library(rethinking)
library(dagitty)

# functions
sim_weight <- function(lengths, sd, b) {
  unobserved <- rnorm(length(lengths), 0, 1) # random noise
  weights <- b*lengths + unobserved
  return(weights)
}
sim_colour <- function(lengths, sd, b) {
  unobserved <- rnorm(length(lengths), 0, 1) # random noise
  colour <- b*lengths + unobserved
  return(colour)
}
sim_predation <- function(colours, weights, sd, bC, bW) {
  unobserved <- rnorm(length(colours), 0, 1) # random noise
  predation <- bC*colours + bW*weights + unobserved
  return(predation)
}
sim_reproduction <- function(colours, weights, sd, bC, bW) {
  unobserved <- rnorm(length(colours), 0, 1) # random noise
  reproduction <- bC*colours + bW*weights + unobserved
  return(reproduction)
}
```

Our objective today is to revisit DAGS, and how we use statistical models to test if our DAGs are consistent with our data.

“DAGs are used to represent causal relationships within a given system. A DAG consists of a set of nodes (variables) that are connected to each other by edges (arrows). These arrows represent causal relationships between variables, pointing from cause to effect, with causes preceding their effects.” (Arif & MacNeil, 2023) \<- Fantastic paper about DAGS out of our lab! See bottom for reference.

Once we have created a DAG, we can test it against our observed data to check for DAG-data consistency.

What does this mean? A DAG often asserts multiple independencies (variables are considered to be independent of each other) that should be true in our data, if both the DAG and the our data are representative of the system.

In a DAG, a pair of variables can be independent of each other (e.g., Weight is independent of Colour) if there are no paths (i.e., a sequence of nodes and arrows) connecting them. Also, a pair of variables can be conditionally independent.

Conditional independencies emerge from separating the dependency (blocking the path). For our DAG to be considered consistent with out data, all of the independencies & conditional independencies need to hold true in our observed data. We can test for this.

## Fork

When something is a common cause of two variables

```{r}
fork_dag <- dagitty("dag{ Length -> Weight Length -> Colour }")
coordinates(fork_dag) <- list(x=c(Length=0, Weight=1, Colour=0), y=c(Length=0, Weight=0, Colour=1))
drawdag(fork_dag)
```

```{r}
# implied independencies of dag
impliedConditionalIndependencies(fork_dag)

# what do i need to adjust for to infer a total causal effect
adjustmentSets(fork_dag, outcome = "Weight", exposure="Colour", effect="total")
```

What does conditional on mean?. In other statistical terminology it also means ‘adjust for’, for us this mean we need to include this variable in our model, and essentially ignore its effect as we are only care about the effect of ONE variable on another.

```{r}
# number of frogs
n_frog <- 75
# generate some lengths
lengths <- rnorm(n_frog, 0,1)
# Length -> Weight
weights <- sim_weight(lengths, 1, 1)
# Length -> Colour
colours <- sim_colour(lengths, 1, -1)

frog_data <- list(
  weights=weights,
  lengths=lengths,
  colours=colours
)

# test for independence
fork_model <- ulam(
  alist(
    weights ~ dnorm(mu, sigma),
    mu <- alpha + bC*colours + bL*lengths,
    alpha ~ dnorm(0,.2),
    bC ~ dnorm(0, 1),
    bL ~ dnorm(0,.5),
    sigma ~ dexp(1)
  ), data=frog_data, chains=4
)
plot(coeftab(fork_model))

```

```{r}
# Do not include length
fork_model2 <- ulam(
  alist(
    weights ~ dnorm(mu, sigma),
    mu <- alpha + bC*colours,
    alpha ~ dnorm(0,.2),
    bC ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=frog_data, chains=4
)
plot(coeftab(fork_model,fork_model2))
```

## Pipe

Here, our causal assumptions are that:

**Length** causes **Weight** which causes **Reproduction** and

**Length** causes **Colour** which causes **Reproduction**

```{r}
pipe_dag <- dagitty("dag{ Length -> Weight Length -> Colour Colour-> Reproduction Weight->Reproduction}")
coordinates(pipe_dag) <- list(x=c(Length=0, Weight=1, Colour=0, Reproduction=1), y=c(Length=0, Weight=0, Colour=1, Reproduction=1))
drawdag(pipe_dag)
```

```{r}
# implied idependencies of dag
impliedConditionalIndependencies(pipe_dag)
```

```{r}
# colour -> reproduction, weight->reproduction
reproduction <- sim_reproduction(colours, weights, sd=1, bC=-1.5, bW=1)

# add reproduction to data set
frog_data$reproduction <- reproduction

# testing for independence
pipe_model <- ulam(
  alist(
    reproduction ~ dnorm(mu, sigma),
    mu <- alpha + bC*colours + bL*lengths + bW*weights,
    alpha ~ dnorm(0, 0.2),
    bC ~ dnorm(0,1),
    bL ~ dnorm(0, 0.8),
    bW ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=frog_data, chains=4
)
precis(pipe_model)
plot(coeftab(pipe_model))
```

```{r}
# what do i need to adjust for to infer a total causal effect
adjustmentSets(pipe_dag, exposure="Colour", outcome = "Reproduction", effect="total")
```

```{r}


# pipe without adjustment
pipe_model_no_adjustment <- ulam(
  alist(
    reproduction ~ dnorm(mu, sigma),
    mu <- alpha + bC*colours,
    alpha ~ dnorm(0,0.2),
    bC ~ dnorm(0,1),
    sigma ~ dexp(1)
  ), data=frog_data, chains=4
)
# pipe with adjustment
pipe_model_w_adjustment <- ulam(
  alist(
    reproduction ~ dnorm(mu, sigma),
    mu <- alpha + bC*colours + bL*lengths,
    alpha ~ dnorm(0,0.2),
    bC ~ dnorm(0,1),
    bL ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ), data=frog_data, chains=4
)
plot(coeftab(pipe_model_no_adjustment, pipe_model_w_adjustment))
```

## Collider

When a variable has two separate causes. Adjusting for colliders opens back door paths

```{r}
collider_dag <- dagitty("dag{ Length -> Weight Length -> Colour Colour-> Reproduction Colour->Predation Weight->Predation Weight->Reproduction}")
coordinates(collider_dag) <- list(x=c(Length=0, Weight=2, Colour=0, Reproduction=2, Predation=1), y=c(Length=0, Weight=0, Colour=2, Reproduction=2, Predation=1))
drawdag(collider_dag)
```

```{r}
impliedConditionalIndependencies(collider_dag)
```

### Simulate our data.

```{r}
# colour -> predation, weight -> predation
predation <- sim_predation(colours, weights, sd=1, bC=1.5, bW=1)

# add predation to data set
frog_data$predation=predation
```

```{r}
# Lngt _||_ Prdt | Colr, Wght
predation_model <- ulam(
  alist(
    predation ~ dnorm(mu, sigma),
    mu <- alpha + bC*colours + bL*lengths + bW*weights,
    alpha ~ dnorm(0,0.2),
    bC ~ dnorm(0,1),
    bL ~ dnorm(0,0.2),
    bW ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ), data=frog_data, chains=4
)
precis(predation_model)
plot(coeftab(predation_model))
```

```{r}
# Lngt _||_ Rprd | Colr, Wght
predation_model2 <- ulam(
  alist(
    predation ~ dnorm(mu, sigma),
    mu <- alpha + bC*colours + bR*reproduction + bW*weights,
    alpha ~ dnorm(0,1),
    bC ~ dnorm(0,1),
    bR ~ dnorm(0,1),
    bW ~ dnorm(0,1),
    sigma ~ dexp(1)
  ), data=frog_data, chains=4
)
precis(predation_model2)
plot(coeftab(predation_model2))
```

References 1. Arif S, MacNeil MA. Applying the structural causal model framework for observational causal inference in ecology. Ecological Monographs. 2023;93(1):e1554. 2. Mcelreath R. Statistical Rethinking 2 A Bayesian Course with Examples in R and Stan Second Edition. 2nd ed. 2020.
